### 2022-02-21 tex/esop22/rebuttal.md

Rebuttal
========

Thank you to the reviewers for their careful reviews and kind comments
and suggestions. We have commented below on the major questions and
concerns brought up by the reviews, and given our suggestions for how
we will amend in the paper for any further revision.

Reviewer A
----------

> Please clarify the "Page 9: As far as I can tell, the *c and I*c and
> r.c used in the definition of [[_]]p are not defined defined
> anywhere. " to help me understand this definition.

These are liftings of the corresponding bunched connectives from page
5 to be indexed by vector/context pairs. We will make this clear in
any revised version. The $\dot\times$ product works because it is a
polymorphic pointwise use of $\times$ for any indexing set. Our
notation on page 5 implies that it is only defined for vectors, but it
is actually polymorphic. We will state this explicitly.

> It might be worth giving some general guidance about how to
> think about the orientation of <=

We think of the ordering `x <= y` as convertibility; so that something
labelled with `x` is sufficient to supply a demand of `y`. This gives
the upside-down ordering when compared the standard natural
numbers. Abel and Bernardy (ICFP 2020) make the same choice.

> In the "An element of a chosen posemiring" paragraph, it would help
> to define <= on vectors (the green scripty P's and Q's that occur in
> the typing rules) --- I assume this means pointwise <= of resources?

Yes, this is the pointwise ordering.

> It would have helped me to have an explicit comment about what
> isomorphisms you're eliding when representing resource vectors (the
> script P and script Q) in different ways.

Conceptually, contexts with usage annotations are pairs of vectors,
one of semiring values and one of types. We have found it useful in
the formalisation to store these vectors in an "unnormalised" fashion
as trees that record their history of appends used to construct
them. This non-canonicity does not cause a problem because we can
always apply renaming to shift to a different association
structure. The isomorphisms come from renamings. We never compare
contexts for equality directly, we only make demands on what variables
appear in them and what their usage annotations and types are. This is
a minor formalisation "trick" that is difficult to get across in a
paper intended for a general audience.

> Page 5, bottom: As far as I understand, the resource 0 doesn't have
> any special status with respect to the preorder <=. So why does R <=
> 0 get a "connective" in the framework

The vector 0 is special with respect to the vector space structure,
being the unit for the addition. There is no multiplication structure
between vectors, and so no use for the vector of all 1s. I* is also
defined with `R <= 0` to ensure it is the unit of the * connective.

> In Definition 5.1, I found the role of the Q' confusing at
> first

You are correct about what it is used for -- it is a convenience. We
will add a comment to clarify.

> [Agda notation]

We will track these down and add the necessary definitions.

> It would be good to give some empirical sense for how well this
> resource elaborator works --- what examples is it able to handle,
> what times out, etc.?

We will elaborate on the elaborator, as also suggested by the other
reviewers.

> The type systems for Amortized Automated Resource Analysis by
> Hofman, Jost, Hoffmann, etc. have the flavor of these
> resource-annotated type systems -- can you encode those?

We can't definitively say, but we think the answer is that we cannot
directly encode these systems because the contraction rule they use
does not fully preserve the types: a variable of type A can be
duplicated to two variables of types A1 and A2 whose integrated
potentials add up to the ones in the type A. In our system,
duplicating a variable affects its top-level usage annotation, but
preserves its type. It may be possible to encode a system for
amortised reasoning however.

> It would also be nice to show that you can encode a system with
> "non-logical" rules, e.g. some of the primitives for differential
> privacy in [18].

This is certainly possible. For the Reed and Pierce system, we would
need to add rules for the graded differentially private distribution
monad, which in our system is just some more rules.

> What are the prospects for extending the framework beyond simple
> types, e.g. to handle resource polymorphism (quantification over
> resources) or dependent types?

We think that extension to polymorphism (over types or usage
annotations) would be relatively straightforward conceptually given
that the kinding or resourcing contexts would not interact with the
substructural parts. It would become harder to make a generic
elaborator, however, due to the need to reason about open terms in the
semiring (something that the Granule systems needs to use external
solvers for). Extension to dependent types is much harder; though we
think that our insights will help extend any formalisation of non
substructural dependent types to having semiring resource annotations.

Reviewer B
----------

> Can only really be followed by an expert in this kind of extremely
> general formalization in Agda.

We have tried to mitigate this by using examples throughout. For a
revised final version we will guide our examples by trying to get
across how the framework is used in practice. For example, an example
of using the elaborator defined in Section 7.3, as suggested.

> Extremely dense in notation and fairly sparse in giving intuitions
> of what's going on.

In addressing reviewer C's points, we will add more intuition to help
readers used to extrinsic presentations of type systems in
Section 3. We will simplify Section 4, as explained below. We will
also more high-level discussion in the later sections on what is
possible with the syntactic traversals for generic metatheory, again
see below. To make space will we move some of the technical content of
Section 5 to an appendix to concentrate more on the high level ideas
of using linear algebra.

> Explanation after page 8 (construction of terms) is too opaque

We can simplify this section by removing the I--OpenFam and I--OpenExt
definitions, since I is always Ty. We will then add proper
explanations for the remaining three defnitions (OpenType, OpenFam and
OpenExtFam) and point to the places where they are required in our
development. The more general definitions are not actually used until
the `extend` function on page 18, where we will explicitly say that we
need the extra generality in a revised version.

> How does one use this for *metatheory* as opposed just creating
> syntaxes.
and
> What kind of metatheory can you do?

We can define functions such as generic renaming and substitution, and
generic denotational semantics. But we cannot yet (easily) reason
about them beyond their types. We expect that the `Simulation`
construct from the Allais et al. paper [3] will smoothly transfer to
this setting, which allow us to prove equational properties of these
traversals. For example, the substitution of a composition of two
simultaneous substitutions being the composition of two separate
subsitutions. In this paper, we wanted to show that the core of Allais
et al. transfers to the semiring-annotated setting, where even
defining the syntax and the operations of renaming and substitution
are nontrivial and require a linearly algebraic approach. We leave
transfer of the equational reasoning principles to future work. We
expect that given our foundational work, equational reasoning will
follow easily.

> what kind of "things" can I expect to reasonably define as
> traversals?
and
> The applicability of the framework in terms of metareasoning beyond
> the use cases in the paper is not really discussed and the
> automation/elaborator is not really exercised in the presentation.

Traversals are a "nicer" fold / recursion principle over the syntax,
that automatically handles dealing with updating of parameters when
going under binders. This is particularly useful in the renaming and
substitution cases, where the parameter is the renaming or
substitution. Using a traversal saves on repeated work that would be
required if one were to do this directly with the "native" recursion
principle for the type of Terms. The key point is that defining
renaming and substitution for substructural systems is not trivial,
and our framework yields a generic definition for a wide range of
such systems.

> An elaborator for this is great, but how can I use it?

We will give an actual use of the elaborator to show how it is used in
the context of Examples 8-10 (the monotone subtraction function),
which will also be compressed to give more space to discussion of the
framework.

> Concretely, I think some of the examples in the paper (e.g. Examples
> 8-10) could be cut for extra space so that definitions, roadmaps and
> actual explanations can be given in a way that makes the paper
> understandable to a wider audience.

We take the reviewer's point that examples 8-10 are more on an example
application of one part of the metatheory (a denotational semantics)
than the overall picture. We will do as the reviewer suggests to
compress these examples to give more space for high level discussion
of the features of the framework. In particular, we will add a short
discussion of how to do denotational semantics for other systems
within the framework. The main point of this section is that our
framework provides a way to formalise much of Abel and Bernardy's ICFP
2020 paper, including the free theorems arising from the usage
annotations.

Reviewer C
----------

> Basically: Expand section 3, starting with a non-intrinsic version
> of Figure 1; explain how to get from the non-intrinsic system to
> Figure 1; explain how to get to Figure 2.

Good idea. We agree that Section 3 is the key section to understanding
our approach. Our plan is to expand the discussion of the intrinsic
presentation in Figure 1 by comparing them to the extrinsic
presentations directly. Hopefully this will lead the reader more
gently to the final presentation in Fig 2. As suggested, comparing our
approach to HOAS's use of ambient contexts is also a useful way to
help readers' intuitions.

> missing some work that may be related: ambient contexts are used in
> higher-order abstract syntax (as in LF/Twelf and its successors)

We will add a discussion of the use of ambient contexts in HOAS
presentations. The interesting point there, we think, is that to
easily encode substructural systems, the common approach is to
replicate the substructural discipline in the logical framework (e.g.,
Linear LF). Here, we are exploiting the same idea internally in Agda
by using the bunched connectives, which means that we don't need to
switch to a completely new logical framework. Thank you for the prompt
to discuss this related work.

> In Definition 2, in the first line, I take it that → is implication,

Yes, this is implication. We will write this out using English words
instead.

> Is the "Rγ" notation chosen for a reason, compared to something like
> R; γ?

We felt that this notation was more concise. Figure 1 with $(R; \gamma)$
everywhere would be much messier, in our opinion.

### 2022-03-18 tex/examples.tex
[ backport ] examples from ESOP22 to thesis

@@

We will take semantics of this system into
\emph{world-indexed relations}~\cite{AbelBernardy2020,context-constrained}.
A world-indexed relation over a poset of worlds \AgdaBound{W} is a set over
which
we have a \AgdaBound{W}-indexed binary relation satisfying a presheaf-like
property with respect to the order on \AgdaBound{W}.

Morphisms between world-indexed relations \AgdaBound{R} and \AgdaBound{S}
consist of a mapping between the underlying sets such that that mapping
preserves relatedness from \AgdaBound{R} to \AgdaBound{S}.

%\ExecuteMetaData[\WReltex]{IR}
%\ExecuteMetaData[\WReltex]{tensorR}
%\ExecuteMetaData[\WReltex]{lollyR}

The final piece of sematics we need is a \emph{bang} operator.
We allow the
semantic \emph{bang} to be an arbitrary annotation-indexed functor on
world-indexed relations.
This functor must respect all of the structure on the indices, making it a
graded comonad over multiplication, as well as being lax monoidal at any
particular index \AgdaBound{r}.

To associate semantics to syntax, we start as standard by associating
world-indexed relations to types.
We also extend the semantics of types to contexts, using \AgdaFunction{I$^R$},
\AgdaFunction{$\otimes^R$}, and \AgdaField{!$^R$} to interpret the empty
context, concatenation, and usage annotations on singletons, respectively.

\ExecuteMetaData[\WReltex]{sem}

The semantics of a term is then to be a morphism from the interpretation of the
context to the interpretation of the term's type.

\ExecuteMetaData[\WReltex]{sem-vdash}

Variables are given semantics by \AgdaFunction{lookup$^R$} (definition omitted).

Now, we give a \AgdaRecord{Semantics}.
The choice of \AgdaBound{$\V$} as
\AgdaRecord{\AgdaUnderscore{}$\sqni$\AgdaUnderscore{}} is somewhat arbitrary,
given that a standard denotational semantics would not use intermediate
environments in the same sense as renaming and substitution, but allows us to
reuse the standard facts that variables support renaming and identity
environments.
With this choice of \AgdaBound{$\V$} and \AgdaBound{$\C$}, we interpret
environment entries by \AgdaFunction{lookup$^R$}.
Meanwhile, for the logical rules, we ignore environments by using
\AgdaFunction{reify} to just deal with morphisms in an extended context.
As such, $\lambda$-abstractions are easy to interpret, while applications
require some massaging to remove the extension by an empty context, followed by
some plumbing to split the interpretation of the context according to the usage
constraints and feed the interpretation of the argument \AgdaBound{n$'$} into
the interpretation of the function \AgdaBound{m$'$}.

In order to map open terms to interpretations, we take the action of the
semantics and give the identity renaming as the starting environment.

\begin{example}
  We can make a subtraction function from primitive addition and negation on
  integers.
  Subtraction is covariant in its first argument and contravariant in its
  second argument.
  We give the definition in pseudocode, as we have not yet seen how to
  conveniently write terms (\cref{sec:usage-elaborator}).

@@

Using the constructs we have seen so far, producing example terms soon becomes
extremely tedious.
We achieved some abbreviation by using pattern synonyms, but we still have to
produce essentially bespoke proofs whenever we use a usage-sensitive part of the
syntax.
The size of each of these proofs is roughly proportional to the number of free
variables, so the amount of proof we have to write grows roughly quadratically
with the size of terms.
An additional factor, which we can't see on paper, is that type checking time
for these proofs soon becomes prohibitive to interactive development.

@@

To actually use \AgdaFunction{elab-sem} on terms, we take the associated
\AgdaFunction{semantics} and pass it the identity environment (an identity
\emph{renaming} in this case, because $\V$ is a family of variables).
\todo{Previous examples are now deleted.}
In the previous examples, we have used \AgdaFunction{elab-unique}, which further
checks statically that exactly one derivation is returned.
The candidate usage vector \AgdaBound{R} will be empty for closed terms, and
otherwise we have to supply the intended usage annotations.

>>>>

In this section, we provide three example uses of semantic traversals: generic
renaming and substitution, a usage elaborator, and a denotational semantics.
The reader is also encouraged to see the far greater range of examples in the
work of \citet{AACMM21}, which should adapt to our usage-annotated setting.
Renaming and substitution are essential results, while the latter two examples
focus on usage annotations.

A result we will use throughout this section is \emph{reification}.
When we have an index-preserving mapping from usage-checked variables to
$\V$-environments, we can construct environments of the form
$\Gamma \env\V \Gamma$ (identity environments) for all $\Gamma$.
This lets us write the \AgdaFunction{reify} function, which  simplifies our
obligations in giving a \AgdaRecord{Semantics} by coercing
\AgdaFunction{Kripke} functions into just
\AgdaBound{$\C$}-values in an extended context.

%To show this, we instantiate the Kripke function with the renaming
%$\swarrow^r : \Gamma, \gr0\delta \env\sqni \Gamma$ to extend the scope, and
%pass it an argument environment
%$\plr{\id^\V; \searrow^r} : \gr0\gamma, \Delta \env\V \Delta$ to fill in the
%extended part.
%Post-composition of a renaming onto an environment comes from
%\cref{thm:env-postren}.

%\ExecuteMetaData[\Syntactictex]{reify}

\begin{lemma}[\AgdaFunction{reify}]\label{thm:reify}
  If $\V$ is an open family such that there is a function
  $v : \forallb{{\sqni} \dotto \V}$, we get a function of type
  $\forallb{\mathrm{Kripke}\,\V\,\C \dotto \mathrm{Scope}\,\C}$ for any $\C$.
\end{lemma}
\begin{proof}
  Let $b : \mathrm{Kripke}\,\V\,\C\,\Delta\,\Gamma\,A$.
  That is, $b$ is a Kripke function yielding $\C$-computations
  We want to apply $b$ so as to get a $\C\,\plr{\Gamma, \Delta}\,A$.
  Let $\grP\gamma = \Gamma$ and $\grQ\delta = \Delta$.
  The $\Box^r$ in the type of $b$ allows us to reverse-rename $\Gamma$ to
  $\Gamma, \gr0\delta$.
  Then we give the $\wand$-function an argument in context
  $\gr0\gamma, \Delta$, noting that
  $\plr{\Gamma, \gr0\delta} + \plr{\gr0\gamma, \Delta} = \plr{\Gamma, \Delta}$,
  as we wanted from the result.
  The argument needs type $\gr0\gamma, \Delta \env\V \Delta$.
  We produce this via \cref{thm:env-postren} from an environment
  $\rho : \gr0\gamma, \Delta \env\V \gr0\gamma, \Delta$ created using $v$
  and a renaming which is the complement to that used on $\Box^r$.
\end{proof}

All of the $\V$s used in examples in this paper support identity environments.
However, \citet[p.~27]{AACMM21} give some important examples that do not
support identity environments, and thus cannot use \AgdaFunction{reify}
(\cref{thm:reify}).
The feature that causes the lack of support for identity environments is that
a semantics can make use of the fact that only variables of particular kinds
are bound by the syntax.
In the examples of \citeauthor{AACMM21}, a bidirectionally typed language only
binds variables that synthesise their type, as opposed to those whose type is
checked.
The semantics of type-checking and elaboration rely on variables synthesising
their type, so \AgdaBound{$\V$} is chosen to cover only those variables.
Instead of using \AgdaFunction{reify}, we must observe that each syntactic form
only binds such synthesising variables.
Similar phenomena would appear in, say, a call-by-value language where
variables are values (not computations), or a polarised language where
variables always have a polarity matching their type.

@@

@@

Using the constructs we have seen so far, producing example terms soon becomes
extremely tedious.
We can achieve some abbreviation by using pattern synonyms to wrap around
\AgdaInductiveConstructor{`con} expressions, but we still have to
produce essentially bespoke proofs whenever we use a usage-sensitive part of the
syntax.
The size of each of these proofs is roughly proportional to the number of free
variables, so the amount of proof we have to write grows roughly quadratically
with the size of terms.
An additional factor, which we can't see on paper, is that type checking time
for these proofs soon becomes prohibitive to interactive development.

@@

To actually use \AgdaFunction{elab-sem} on terms, we take the associated
\AgdaFunction{semantics} and pass it the identity environment (an identity
\emph{renaming} in this case, because $\V$ is a family of variables).
We use \AgdaFunction{elab-unique}, which further
checks statically that exactly one derivation is returned.
The candidate usage vector \AgdaBound{R} will be \AgdaFunction{[]} for closed
terms, and otherwise we have to supply the intended usage annotations.

We can now use the elaborator to automatically infer the usage
annotations for the example at the end of \cref{sec:terms}. This
allows us to write:
\ExecuteMetaData[\PaperExamplestex]{cojoin}
We have instantiated the usage elaborator so that:
\AgdaField{0\#$^{-1}$} is a singleton on $\gr0$ and $\gr\omega$, and
empty on $\gr1$; \AgdaField{1\#$^{-1}$} is a singleton on $\gr1$ and
$\gr\omega$, and empty on $\gr0$; \AgdaField{+$^{-1}$} gives $\gr0
\mapsto [(\gr0,\gr0)]$, $\gr1 \mapsto [(\gr0,\gr1),(\gr1,\gr0)]$, and
$\gr\omega \mapsto [(\gr\omega,\gr\omega)]$; and \AgdaField{*$^{-1}$}
gives $(\gr\omega, \gr0) \mapsto [\gr0]$, $(\gr\omega, \gr1) \mapsto
[]$, and $(\gr\omega, \gr\omega) \mapsto [\gr\omega]$ (omitting
$(\gr0, \_)$ and $(\gr1, \_)$ cases for brevity). Note that we do not
consider splitting $\gr\omega$ up as, say, $\gr1 + \gr\omega$, because
this splitting would introduce more non-determinism but not allow any
more terms to be typed. As such, the only non-determinism comes when
we have variables annotated $\gr1$ and need to do an additive split,
like when we apply the \AgdaInductiveConstructor{!E} rule below. At
this point, the variable can become either $\gr0$-annotated in the
left subterm and $\gr1$-annotated on the right, or vice-versa. We will
find that, because the left subterm wants to use that variable, the
former choice will be rejected. The function \AgdaFunction{var\#} is a
convenience for converting statically known natural numbers,
representing de Bruijn \emph{levels}, into variable terms.

We will take semantics of this system into
\emph{world-indexed relations}~\cite{AbelBernardy2020,context-constrained}.
A world-indexed relation (\AgdaRecord{WRel}) over a poset of worlds
\AgdaBound{W} is a set over which
we have a \AgdaBound{W}-indexed binary relation satisfying a presheaf-like
property with respect to the order on \AgdaBound{W}.

Morphisms (\AgdaRecord{WRelMor}) between world-indexed relations \AgdaBound{R}
and \AgdaBound{S} consist of a mapping between the underlying sets such that, at
each fixed world \AgdaBound{w}, the mapping preserves relatedness from
\AgdaBound{R} to \AgdaBound{S}.

The final piece of semantics we need is a \emph{bang} operator.
We allow the
semantic \emph{bang} to be an arbitrary annotation-indexed functor on
world-indexed relations.
This functor must respect all of the structure on the indices, making it a
graded comonad over multiplication, as well as being lax monoidal at any
particular index \AgdaBound{r}.

%To associate semantics to syntax, we start as standard by associating
%world-indexed relations to types.
%We also extend the semantics of types to contexts, using \AgdaFunction{I$^R$},
%\AgdaFunction{$\otimes^R$}, and \AgdaField{!$^R$} to interpret the empty
%context, concatenation, and usage annotations on singletons, respectively.
%
%\ExecuteMetaData[\WReltex]{sem}
%
%The semantics of a term is then to be a morphism from the interpretation of the
%context to the interpretation of the term's type.
%
%\ExecuteMetaData[\WReltex]{sem-vdash}

The semantics of a type is given by \AgdaFunction{$\llbracket$\_$\rrbracket$},
which maps into world-indexed relations.
The function type is interpreted using \AgdaFunction{$\multimap^R$} and
\AgdaField{!$^R$}.
Contexts are interpreted by \AgdaFunction{$\llbracket$\_$\rrbracket^c$}, using
\AgdaFunction{$\otimes^R$} and \AgdaFunction{I$^R$}.
Terms are interpreted as morphisms by the open family
\AgdaFunction{$\llbracket$\_$\vdash$\_$\rrbracket$}.
Variables are interpreted by \AgdaFunction{lookup$^R$} (definition omitted).

Now we give a \AgdaRecord{Semantics}.
The choice of \AgdaBound{$\V$} as
\AgdaRecord{\AgdaUnderscore{}$\sqni$\AgdaUnderscore{}} is somewhat arbitrary,
given that a standard denotational semantics would not use intermediate
environments in the same sense as renaming and substitution, but it allows us to
reuse the standard facts that variables support renaming and identity
environments.
With this choice of \AgdaBound{$\V$} and \AgdaBound{$\C$}, we interpret
environment entries by \AgdaFunction{lookup$^R$}.
Meanwhile, for the logical rules, we ignore environments by using
\AgdaFunction{reify} to just deal with morphisms in an extended context.
As such, $\lambda$-abstractions are easy to interpret, while applications
require some massaging to remove the extension by an empty context, followed by
some plumbing to split the interpretation of the context according to the usage
constraints and feed the interpretation of the argument \AgdaBound{n$'$} into
the interpretation of the function \AgdaBound{m$'$}.

Then, the semantics of terms is given by the function
\AgdaFunction{semantics}\AgdaSpace{}\AgdaFunction{Wrel}\AgdaSpace{}%
\AgdaFunction{1$^r$}, where \AgdaFunction{1$^r$} is the identity renaming.
%In order to map open terms to interpretations, we take the action of the
%semantics and give the identity renaming as the starting environment.

\begin{example}\label{thm:minus}
  We can make a subtraction function from primitive addition and negation on
  integers.
  Subtraction is covariant in its first argument and contravariant in its
  second argument.
  We give the definition in pseudocode, though it is also amenable to the
  usage elaborator of \cref{sec:usage-elaborator}, suitably instantiated.
